A Recipe for Scavenging - the natural history of a behaviour

### 2016-04-21 natural history of scavenging.tex
More updates

@@

%KH still not sure about this transition.
Perhaps the most extreme example of utilizing such late carcass stage material is the ability to eat bone, a resource that is typically too hard for many predators to process. 
% Thus, some phenotypic adaptations such as the ability to crush bone reveal another useful scavenger trait.
Osteophagy is known across a range of terrestrial carnivores and given that some fat-rich mammalian bones have an energy density (6.7 kJ/g) comparable with that of muscle tissue, it makes skeletal remains an enticing resource \citep{brown1989study}.
This ability reached its zenith among hyenas with the evolution of the estimated 110 kg \textit{Pachycrocuta brevirostris} during the Pliocene \citep[3.6 - 2.58 Mya; ][]{palmqvist2011giant}.
Indeed, their extinction has been blamed on the decline of sabretooth cats (Machairodontinae), the unique skull morphology of the latter meant they would leave a large amount of food on a carcass for would-be scavengers \citep{palmqvist2011giant}. 
Earlier in the evolution of mammals, the bone-crushing dogs that evolved during the Oligocene (Borophaginae; 33.9 - 23.03 Mya) have also been compared to hyenas in terms of their feeding ecology \citep{van2003chapter,martin2016pursuit}.

@@

Interestingly, such comparisons have given insight into the feeding ecology of early hominins who, for instance, had the ability to craft tools for breaking open bones \citep{ARCM:ARCM12084}.
The question of where our ancestors placed on the hunter-scavenger axis during the Plio-Pleistocene has been a matter of debate for years \citep{dominguez2002hunting}.
%%KH I think this shouldn't be the first place we mention humans as they do appear and fit well in other sections.
A recent study investigating potential scavenging opportunities for hominins in Kenya found that, even when discounting bone material, there is a substantial amount of scavengeable meat left on predated remains; sufficient to sustain the requirements of an adult male \textit{Homo erectus} \citep{pobiner2015new}.
In some historical hominin-inhabited areas there were a greater number of felids than hyenids.
Again, this is significant because hyenas are likely to have left far less flesh on a carcass than a felid such as a sabretooth, enabling contemporaneous hominins to benefit \citep{pobiner2015new}.
The use of tools and the cooperative nature of hominins meant they could likely get a substantial part of their energetic requirements through scavenging depending on their environment \citep{moleon2014humans}.
%KH should this go in the prey availability section?

In Mesozoic systems some extremely large theropod dinosaurs had a morphology indicative of an ability to process bone \citep[e.g. the robust skull and dentition of \textit{Tyrannosaurus rex}][]{hone2010feeding}.
There is direct evidence that \textit{T. rex} did this in the form of distinctive wear marks on its tooth apices \citep{farlow1994wear,schubert2005wear} and the presence of bone fragments in its coprolites \citep{chin1998king}.
The animal also had an enormous bite force, with one estimate putting it at 57000 Newtons \citep{bates2012estimating} which would have been powerful enough to break open skeletal material \citep{rayfield2001cranial}.
%KH I would mention that however, our paper shows that such species that would have been able to have such bites were probably too large for this to be an effective stratagy. Then you can segue into the body size thing.

@@

In their work, the authors demonstrated that species of \textit{intermediate} body masses would have gained the most benefit from scavenging \citep{kane2016body}.
This was the result of gut capacity limitations and the effects of competition at the carcass.
At the larger extreme this owes to the fact that gut capacity doesn't scale isometrically with body mass so the benefits of greater mass level off; there's only so much food an individual can consume at a single sitting \citep{calder1996size}.
For the smaller species, larger competitors would have prevented their access to carrion.

@@

%%%%%%%%%%  AK This annoying paragraph's new home TG: well done finding him a home!
Certainly, scavenging should be particularly attractive to flying species compared to mammals. % Not sure about this paragraph 
The latter can kill prey up to the same body mass as themselves and sometimes an order of magnitude heavier \citep[e.g. socially hunting lions; ][]{owen2008predator}.
In contrast, birds of prey tend to kill prey smaller than themselves \citep{slagsvold2007prey} because of the greater cost of injury and the need to carry off their food.
% This is likely due to their need to kill animals that they can fly away with, as well as the risk of injury being higher (which carries a higher mortality risk) for a bird than a mammal.
Scavenging provides a means for birds to exploit species that would otherwise be too big for them to kill.
% Thus, for birds, scavenging means they can exploit species that would otherwise be too big for them to kill.
%%%%%%%%%%this is moved from the locomotion section as it didn't seem to fit.

@@

\section*{Prey Availability}
%Both the biotic and abiotic environment a would-be scavenger finds itself in can influence to degree to which it can depend on carrion. %TG: This sentence is convoluted: maybe go with the following:
The position of a species on the scavenging scale can also be influenced by the availability of carrion in the environment, which is dependent on biotic and abiotic factors.
Aspects including, primary productivity, relief, temperature and competition will all greatly affect scavenging tendency. 
\cite{ruxton2004obligate} suggest a system with a productivity similar to the Serengeti could have supported a mammalian or reptilian terrestrial scavenger.
Indeed, in systems that were dominated by large ectothermic or mesothermic vertebrates, the same primary productivity would have supported a greater biomass \citep{mcnab2009resources}.
%TG: We're now talking about Late Mesozoic systems right? Needs to be mentioned somewhere.
The upshot of this is that there was a higher biomass of herbivores dying and offering scavenging opportunities.
Predators were large-bodied too compared to extant mammalian predators \citep{mcnab2009resources}, and so, especially if they were ectothermic, could last longer between meals, rendering scavenging a more attractive behaviour relative to predation.
Osteophagy may have been even more viable during the Mesozoic era as well because of this skewed body mass distribution of herbivores towards larger sizes \citep{10.1371/journal.pone.0051925}.
When we couple this with the fact that skeletal mass scales greater than linearly with body mass \citep{prange1979scaling} there would have been a lot of bone material to consume in the environment provided an animal had the biology to process it \citep{chure1997one}.
% As we discussed earlier, this ability is often extremely beneficial to a scavenger.

Frequently, the interplay between abiotic and biotic factors can impact the ability of an animal to scavenge. 
We know vultures and eagles tend to soar using thermals and if these air pockets don't form, say on a cloudy day, the bird is grounded \citep{mundy1992vultures}.
In many habitats (e.g. the Arctic) it is simply not possible for sufficiently powerful thermals to form and as a consequence large-bodied vultures cannot exist.
One result of this is that terrestrial carnivores like bears and wolves take more carrion \citep{devault2003scavenging}.
Certainly, a major difficulty for terrestrial scavengers is competition with vultures.
Nocturnal behaviour in the hyaenidae in general has been put forth as an adaptation to reduce competition with these exclusively diurnal birds \citep{gittleman2013carnivore}.
If we apply this line of reasoning over evolutionary time-scales, the absence of flying vertebrates in the Palaeozoic may have permitted terrestrial forms to take in a higher proportion of carrion in their diet.

@@

%TG: maybe change transition here.
Staying in the aquatic setting, the phenomenon of occasional bounties of carrion in the form of whale falls has led some researchers to investigate if a scavenger could survive by seeking out these remains exclusively.
\cite{ruxton2005searching} argued that although this is energetically feasible it's ecologically unlikely.
Any animal that could find such whale carcasses is unlikely to have ignored other types of carrion.
Although no aquatic species have ever exceeded the size of whales, some enormous animals have evolved in this environment before the evolution of cetaceans, including \textit{Leedsichthys}, a bony fish from the Middle Jurassic (174.1-163.5 Mya) and the aquatic Mesozoic reptiles, the plesiosaurs, pliosaurs and ichtyosaurs, that could all exceed 15 metres in length \citep{ruxton2011zoology}.
So, despite being unlikely, the energetic feasibility of a marine scavenger that specialises on large carcasses has a long history.
One point of interest is that of the whaling industry, which provided a bonanza of floating carcasses especially during the 20th century \citep{Whitehead415}.
This meant killer whales could switch from hunting to scavenging, a switch made that much easier by the noise of the whaling vessels that would effectively ring the ``dinner-bells" \citep{Whitehead415}.

@@

>>>>

% KH maybe my issue is that the aquatic, terrestrial, aerial structure, which works well for the locomotion, struggles in the the other sections. It kinda leads to sentences that look like they were dropped in and from the discussion to prematurely end on one point before its fully developed.
% KH an alternative structure would be process type structure that would follow the order of events at the carcass. Something like competition, processing the food and then digestion?

@@

%KH still not sure about this transition.
One adaptation is to utilize late carcass stage material that is less subject to decomposition and unavailable to other competitors, for example bone.
% Thus, some phenotypic adaptations such as the ability to crush bone reveal another useful scavenger trait.
Osteophagy is known across a range of terrestrial carnivores and given that some fat-rich mammalian bones have an energy density (6.7 kJ/g) comparable with that of muscle tissue, it makes skeletal remains an enticing resource \citep{brown1989study}.
This ability reached its zenith among hyenas with the evolution of the estimated 110 kg \textit{Pachycrocuta brevirostris} during the Pliocene \citep[3.6 - 2.58 Mya; ][]{palmqvist2011giant}.
Indeed, their extinction has been blamed on the decline of sabretooth cats (Machairodontinae), the unique skull morphology of the latter meant they would leave a large amount of food on a carcass for would-be scavengers \citep{palmqvist2011giant}. 
Earlier in the evolution of mammals, the bone-crushing dogs that evolved during the Oligocene (Borophaginae; 33.9 - 23.03 Mya) have also been compared to hyenas in terms of their feeding ecology \citep{van2003chapter,martin2016pursuit}.

@@

Interestingly, such comparisons have given insight into the feeding ecology of early hominins who, for instance, had the ability to craft tools for breaking open bones \citep{ARCM:ARCM12084}.
The question of where our ancestors placed on the hunter-scavenger axis during the Plio-Pleistocene has been a matter of debate for years \citep{dominguez2002hunting}.
%%KH I think this shouldn't be the first place we mention humans as they do appear and fit well in other sections.
%KH Here and in other section there is a lack of mentioning fossil/remains etc. For example I'm sure there is something about the tools or the marks on the bones that would fit in well here.

A recent study investigating potential scavenging opportunities for hominins in Kenya found that, even when discounting bone material, there is a substantial amount of scavengeable meat left on predated remains; sufficient to sustain the requirements of an adult male \textit{Homo erectus} \citep{pobiner2015new}.
In some historical hominin-inhabited areas there were a greater number of felids than hyenids.
Again, this is significant because hyenas are likely to have left far less flesh on a carcass than a felid such as a sabretooth, enabling contemporaneous hominins to benefit \citep{pobiner2015new}.
The use of tools and the cooperative nature of hominins meant they could likely get a substantial part of their energetic requirements through scavenging depending on their environment \citep{moleon2014humans}.
%KH should this go in the prey availability section?

In Mesozoic systems some extremely large theropod dinosaurs had a morphology indicative of an ability to process bone \citep[e.g. the robust skull and dentition of \textit{Tyrannosaurus rex}][]{hone2010feeding}.
There is direct evidence that \textit{T. rex} did this in the form of distinctive wear marks on its tooth apices \citep{farlow1994wear,schubert2005wear} and the presence of bone fragments in its coprolites \citep{chin1998king}.
The animal also had an enormous bite force, with one estimate putting it at 57000 Newtons \citep{bates2012estimating} which would have been powerful enough to break open skeletal material \citep{rayfield2001cranial}.
%KH something on the robust morphology and size required to generate such forces however essentially restricts flying species from having such a strategy, unless they can use tools (dosent some bird fly up and then drop bones.)
%KH I would also mention that however, our paper shows that such species that would have been able to have such bites were probably too large for this to be an effective stratagy. Then you can segue into the body size thing.
%%An alternative structure would be to go straight into morphologies that crack bones, i.e large heads, then follow such morphology restricts some groups for example flying species or smaller animals like humans. Then you could move into tool use and similar type behavior in humans and birds.

@@

In their work, the authors demonstrated that species of \textit{intermediate} body masses would have gained the most benefit from scavenging \citep{kane2016body}.
This was the result of gut capacity limitations and the effects of competition at the carcass.
% KH I would mention that competition really only affects smaller individuals in this scenario. That is without competition in the model big things did about just as well but small ones did much better. Maybe take the shift in optimum between the models as an example of the affect. 
At the larger extreme this owes to the fact that gut capacity doesn't scale isometrically with body mass so the benefits of greater mass level off; there's only so much food an individual can consume at a single sitting \citep{calder1996size}.
%% KH This sound specific to theropods while if I remember correctly its for mammals. I think we may need to say this in a more general aspect (again changing structure may help. (i.e a stomach section would allow snakes to be included and other extremes))
For the smaller species, larger competitors would have prevented their access to carrion.

@@

%KH Just as a though, people bring this up a lot that body size allows for long period of starvation being specific for scavenging but I never seen anything past a verbal argument. Is there anything that really supports it that say locomotion cant explain (i.e large search area thing)?

%%%%%%%%%%  AK This annoying paragraph's new home TG: well done finding him a home!
%% KH Does it need any home? 
Certainly, scavenging should be particularly attractive to flying species compared to mammals. % Not sure about this paragraph 
The latter can kill prey up to the same body mass as themselves and sometimes an order of magnitude heavier \citep[e.g. socially hunting lions; ][]{owen2008predator}.
In contrast, birds of prey tend to kill prey smaller than themselves \citep{slagsvold2007prey} because of the greater cost of injury and the need to carry off their food.
% This is likely due to their need to kill animals that they can fly away with, as well as the risk of injury being higher (which carries a higher mortality risk) for a bird than a mammal.
Scavenging provides a means for birds to exploit species that would otherwise be too big for them to kill.
% Thus, for birds, scavenging means they can exploit species that would otherwise be too big for them to kill.
%%%%%%%%%%this is moved from the locomotion section as it didn't seem to fit.

@@

\section*{Prey Availability}
%Both the biotic and abiotic environment a would-be scavenger finds itself in can influence to degree to which it can depend on carrion. %TG: This sentence is convoluted: maybe go with the following:
The position of a species on the scavenging scale can also be influenced by the availability of carrion in the environment, which is dependent on biotic and abiotic factors.
Aspects including, primary productivity, relief, temperature and competition will all greatly affect scavenging tendency. 
\cite{ruxton2004obligate} suggest a system with a productivity similar to the Serengeti could have supported an obligate mammalian or reptilian terrestrial scavenger.
%% KH Be careful about slipping between scavenger and obligate scavenger.
Indeed, in systems that were dominated by large ectothermic or mesothermic herbivore vertebrates, the same primary productivity would have supported a greater biomass, due to the scaling of mass with metabolic rate \citep{mcnab2009resources}.
%KH Need to make it clear why here. Only works for larger herbivores as well
%TG: We're now talking about Late Mesozoic systems right? Needs to be mentioned somewhere.
The upshot of this is that there was a higher biomass of herbivores dying and offering scavenging opportunities.
%KH  I have a feeling the population dynamics to this are actually pretty complex. Large animals may live longer and represent a standing biomass similar to tropical rainforest's were all the nutrients are locked up in big trees. The life-history of these animals would hence be what really determines the production of carcasses however that is way outside the scope of this paper and would require a paper all on its own. Maybe we could drop in that point as I have not seen anyone ever mention this point.
Predators were large-bodied too compared to extant mammalian predators \citep{mcnab2009resources}, and so, especially if they were ectothermic, could last longer between meals, rendering scavenging a more attractive behaviour relative to predation.
%%KH Not sure this sentence belongs here, already covered int he handling section.
Osteophagy may have been even more viable during the Mesozoic era as well because of this skewed body mass distribution of herbivores towards larger sizes \citep{10.1371/journal.pone.0051925}.
When we couple this with the fact that skeletal mass scales greater than linearly with body mass \citep{prange1979scaling} there would have been a lot of bone material to consume in the environment provided an animal had the biology to process it \citep{chure1997one}.
% As we discussed earlier, this ability is often extremely beneficial to a scavenger.

Frequently, the interplay between abiotic and biotic factors can impact the ability of an animal to scavenge. 
We know vultures and eagles tend to soar using thermals and if these air pockets don't form, say on a cloudy day, the bird is grounded \citep{mundy1992vultures}.
In many habitats (e.g. the Arctic) it is simply not possible for sufficiently powerful thermals to form and as a consequence large-bodied vultures cannot exist.
%% KH I dont think this belong in this section, we probably have covered this enough in the locomotion part.
One result of this is that terrestrial carnivores like bears and wolves take more carrion \citep{devault2003scavenging}.
Certainly, a major difficulty for terrestrial scavengers is competition with vultures.
Nocturnal behaviour in the hyaenidae in general has been put forth as an adaptation to reduce competition with these exclusively diurnal birds \citep{gittleman2013carnivore}.
If we apply this line of reasoning over evolutionary time-scales, the absence of flying vertebrates in the Palaeozoic may have permitted terrestrial forms to take in a higher proportion of carrion in their diet.
%%KH I think the focus needs to be more on the actual environment and opportunities available. I would suggest that compititon either stays in the handling section (my preference) or is moved down to here (which also works really). 
%%KH You could then split the structure into production in terrestial/marine which focuses on primary productivity and species composition and how they affect what available then the next part being competition in both section.

@@

%TG: maybe change transition here.
Staying in the aquatic setting, the phenomenon of occasional bounties of carrion in the form of whale falls has led some researchers to investigate if an obligate scavenger could survive by seeking out these remains exclusively.
\cite{ruxton2005searching} argued that although this is energetically feasible it's ecologically unlikely.
Any animal that could find such whale carcasses is unlikely to have ignored other types of carrion.
Although no aquatic species have ever exceeded the size of whales, some enormous animals have evolved in this environment before the evolution of cetaceans, including \textit{Leedsichthys}, a bony fish from the Middle Jurassic (174.1-163.5 Mya) and the aquatic Mesozoic reptiles, the plesiosaurs, pliosaurs and ichtyosaurs, that could all exceed 15 metres in length \citep{ruxton2011zoology} (reptile fall ref here).

%% KH check out this paper on Mesozoic reptile falls http://www.nature.com/ncomms/2014/140910/ncomms5789/full/ncomms5789.html
So, despite being unlikely, the energetic feasibility of a marine scavenger that specialises on large carcasses has a long history.
One point of interest is that of the whaling industry, which provided a bonanza of floating carcasses especially during the 20th century \citep{Whitehead415}.
This meant killer whales could switch from hunting to scavenging, a switch made that much easier by the noise of the whaling vessels that would effectively ring the ``dinner-bells" \citep{Whitehead415}.

@@

% KH Maybe we need to more strongly reaffirm that the framework that we give really is useful and should be applied to all sorts of things. etc
